Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-13 rebuttal.tex

\comment{P5, c2, l42: formal citation for one of Hopfinger?s papers is
  required here.}
\response{The relevant reference has been added}

\comment{P5, c2, l58: this assertion is made with no evidence to back
  it up. Why should multi-target models be more reliable? It is not
  clear to me that they should.}\marginpar{\textbf{Needs more work}}
\response{We have rephrased this statement to be less dogmatic}

\comment{P6, c1, l7: I think it is important for the authors to cover
  ELNs by using actual recorded data. Do we know that many scientists
  changed to ELNs ?many years ago? as stated? Is there a reference for
  this or data to back it up?}
\response{We have expanded our original wording, and we now cite two
  references about the use and growth of ELNs in chemistry and
  elsewhere.}

\comment{P6, c1, l7: the authors do not mention the leading players in
  ELNs (e.g. CambridgeSoft, Accelrys, etc.) \ldots I think the
  key software houses that develop and supply these ELN systems should
  be mentioned. Furthermore, systems such as Reaxys should be
  mentioned \ldots} 
 \response{We now mention CambridgeSoft, Accelrys and
  Reaxys in the article, and provide a citation to a review of 35 ELNs
  that was published in November 2011.}

### 2011-11-13 rebuttal.tex

\comment{RDKit is not mentioned at all in this article and is widely
  used as a cheminformatics toolkit in industry and academia and free
  and open-source (although from commercial beginnings).}
\response{We agree that this toolkit should have been included. We
  have updated the text on page 7 to take note of RDKit, though we
  still primarily focus on the CDK and OpenBabel. However, Table 2
  includes RDKit in the feature comparison}

\comment{The authors could more appropriately cover the graph theory
  aspects of molecular structures by correctly applying the
  terminology to which computer scientists and cheminformaticians will
  most certainly be aware}
\response{The text has been updated}

\comment{The inclusion of ontologies is not really useful here as the
  conclusions drawn are vague and do not really inform the current
  state-of-the-art and usefulness of ontologies in cheminformatics.}
\response{The tetx has been updated on page 4 to make the discussion
  of ontologies more focused, specifically identfying challenges in
  linking chemical information to other data types and how ontologies
  and semantic methods can be used ot address this.}

\comment{Section 2.2. Structure Enumeration: it is not clear what this
  section is trying to convey to the reader \ldots}
\response{We have rewritten this section to simplify the language as
  well as be more explicit in applications of and challenges in this
  technique. We have also provide relevant references to key topics in
  this area}

### 2011-11-13 paper.tex
redid the structure enumeration section

To illustrate the difficulty of standardizing chemical structures,
different representations of the same chemical structure are shown in
Figure~\ref{figure:chemical-structures}. The representations range
from a graph (Fig.~\ref{figure:chemical-structures}A) that contains
much implicit information (such as unlabeled nodes represent carbons
with valencies satisified by hydrogens) to a fully labeled graph,
where all information is explicit
(Fig.~\ref{figure:chemical-structures}B). The former is what is
usually exchanged informally between chemists, but to actually mine
chemical structures computationally, something like the latter is
required.  For example, graph mining algorithms can be applied to
labeled graph representations, to discover patterns in large-scale
datasets \cite{wegner2006,horst2009}.

@@

% Fingerprints also facilitate the rapid calculation of quantitative
% structure-based similarity measures. An example measure of
% similarity is the \emph{Tanimoto coefficient}, calculated as the
% ratio $T(a,b) = \frac{c}{(a+b-c)}$, where $c$ is the count of bits
% on in the same position in both the two fingerprints, $a$ is the
% count of bits on in the first structure, and $b$ is the count of
% bits on in the second structure. Historically, this coefficient
% originated from the comparison of ecosystems around lakes, where a
% measure was needed to describe how similar those ecosystems are with
% respect to the animals, plants etc. they contain. However, it can be
% applied to every other system as well where features can be
% described in a binary manner. The Tanimoto coefficient varies in the
% range 0.0 -- 1.0, with a score of 1.0 indicating that the two
% structures are either very similar or even identical (i.e. their
% fingerprints are the same).
%

% Structure-based similarity is however of limited use in grouping of
% chemicals based on non-structural features, such as shared bioactivity
% profiles. For this purpose, a large-scale and flexible classification
% structure is needed for chemicals and their linked biological
% information. One such answer to this need is provided in the form of
% \emph{ontologies}.  An ontology is a formal specification of entities
% and their relationships in a domain of interest. It is structured
% around an underlying logical representation such as Description
% Logics~\cite{baaderdl2007}. The logic-based representation allows
% automated reasoning to derive new knowledge (inferences) from the
% knowledge encoded, making the representation both compact and
% powerful.  Ontologies are already in widespread use across the
% biomedical domain, and in the cheminformatics field ontologies
% relating to the chemical structures themselves and the protein targets
% the molecules are active against are probably most common. A
% similarity measure based on the information encoded in ontologies is
% termed \emph{semantic similarity}, and such measures have already been
% applied to enhance chemical classification in several problem areas
% \cite{couto2010}. Many open questions still remain about the best way
% to combine the information encoded in chemical structures with
% chemical ontologies \cite{hastingsowled2010}.

\subsection{Structure Enumeration}
\label{sec:struct-enum}

Enumerating molecules is a combinatorial problem that has fascinated
chemists, computer scientists and mathematicians alike, for more than
a century. It is one important risk-reduction element in still being
able to encode closely-related molecules and for keeping the space
complexity of databases under (algorithmic) control. Indeed, while
trying to solve the problem of counting the isomers of paraffin
structures or counting substituted aromatic compounds, fundamental
principles in graph theory and combinatorics were developed by Cayley,
Polya and others. Even terms that are widely-used nowadays, like
``graph'' and ``tree'', were originally coined in a chemistry context
(Chapters 1 and 8 in \cite{faulon2010}). More than four decades ago,
Joshua Lederberg, Carl Djerassi, Edward Feigenbaum and others from
Stanford University started developing codes to explicitly enumerate
molecules. DENDRAL is generally quoted in artificial intelligence
textbooks as a pioneer project that produced the first expert
system~\cite{DENDRAL}.

Enumerating molecules is not only an interesting academic exercise but
has practical utility as well. The foremost application of enumeration
is \emph{structure elucidation} of natural compounds such as
metabolites. Metabolites are small molecules that can be found an
organism (such as a human) and which in many cases indicate the state
the organism is in -- such as indicating a disease, or a fasted state,
etc. -- and with a sufficiently large database the idea is to be able
to detect the chemical composition of body fluids and to arrive at
conclusions regarding the particular state of the organism at this
point in time.  Ideally, the wishful bench chemist collects
experimental data from an unknown compound, the data is fed into a
code, and the resulting structure is returned. Although that
streamlined process may not yield a unique solution, there are
commercial software products such as MOLGEN (\url{http://molgen.de})
that can, for instance, list all structures matching a given molecular
formula, or mass spectrometry (MS) data. Another important application
of enumeration is \emph{molecular design}. Here the problem is to
design compounds (drugs, for instance) that optimize some physical,
chemical, or biological property or activity. Although less prolific
and more recent than structure elucidation, molecular design has
introduced some novel solutions to molecular enumeration. Finally,
with the advent of \emph{combinatorial chemistry} (the rapid synthesis
of different but related molecules), molecular enumeration has taken a
central role as it allows computational chemists to construct virtual
libraries and test hypotheses, and it provides guidance in the design
of optimal combinatorial experiments.

The major difficulty with enumeration (apart from the size of the
search space, which is thought to be in the order of $10^{60}$
molecules for 30 heavy atoms \cite{Bohacek:1996ve}) is that the
\emph{in silico} representation of molecules, the so-called molecular
graphs, are labeled objects (i.e. labeled graphs), while the atoms of
a molecule are of course not uniquely labeled. The mathematical
concept to tackle this problem is to consider orbits of labeled
molecular graphs under the operation of the symmetric group. While the
complexity of molecular graph enumeration remains an open problem it
has nonetheless been shown that molecules can be sampled
efficiently \cite{goldberg1999}. Sampling procedures based on the
Metropolis or the Genetic Algorithms have been developed and used to
elucidate natural compounds from NMR data.

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a billion chemical structures~\cite{GDB}. Yet, the
current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community. In addition, GDB-13 only enumerates
molecules up to 13 heavy atoms -- given the possibly exponential
increase in search space with the number of atoms, it seems unlikely
that the practically relevant chemical space of about 30 heavy atoms
will be exhaustively enumerated in the near future.

Apart from analytical chemistry and compound design, enumeration plays
a role is in the \emph{generation of chemical reaction networks}. The
problem here consists of enumerating all the possible compounds that
can be produced by applying reaction rules to a set of initial
molecules. By reversing the reaction rules one can also find sets of
starting compounds necessary for producing a given target. This latter
process is known in chemistry as \emph{retrosynthesis}, ever since the
works of E.J. Corey, who received the Nobel Prize in Chemistry in
1990. Designing new drugs or chemicals, understanding the kinetics of
combustion and petroleum refining, studying the dynamics of metabolic
networks, applying metabolic engineering and synthetic biology to
produce heterologous compounds in microorganisms, all are applications
that require the enumeration of reaction networks. As reviewed in
Chapter 11 in \cite{faulon2010} several network enumeration techniques
have been developed. However, these techniques generally suffer from a
combinatorial explosion of intermediate compounds being produced. One
way to limit the number of compounds generated is to simulate the
dynamics of the network while it is being constructed and remove
compounds of low concentration. Following that idea, methods have been
developed based on the Gillespie Stochastic Simulation Algorithm (SSA)
to compute on-the-fly species concentrations. Chemical reaction
network enumeration and sampling is an active field of research,
particularly in the context of metabolism, either to study
biodegradation, or to propose metabolic engineering strategies to
biosynthesize compounds of commercial interest. The difficulty with
metabolic network design is that in addition to network generation
based on reactions, one also needs to verify that there are enzymatic
events possible for enabling reactions catalyses. That additional task
requires encompassing both chemical and sequence information and the
development of tools that are at the interface between cheminformatics
and bioinformatics.

@@

The basis of predictive modeling in cheminformatics is that the
biological activity of a molecule is a function of its chemical
structure. Together with the \emph{similarity property principle}
\cite{Johnson:1990qf} (structurally similar molecules will exhibit
similar properties), the goal of any modeling approach is to capture
and characterize the correlations between structural features and the
observed biological activities. Simultaneously, such approaches must
also describe the likelihood of error when using such
models for decision making.  A variety of approaches can be employed
to assess the error in (or conversely, the reliability or confidence of) a
prediction, ranging from statistical approaches (XXX) to more
empirical approaches such as defining an \emph{applicability domain} -
the region of the input (i.e., chemical structures) that can be
reliably predicted, usually deliniated using some form of similarity
to the training set.

>>>>

To illustrate the difficulty of standardizing chemical structures,
different representations of the same chemical structure are shown in
Figure~\ref{figure:chemical-structures}. The representations range
from a graph (Fig.~\ref{figure:chemical-structures}A) that contains
much implicit information (such as unlabeled nodes represent carbons
with valencies satisified by hydrogens) to a fully labeled graph,
where all information is explicit
(Fig.~\ref{figure:chemical-structures}B). The former is what is
usually exchanged informally between chemists, but to actually mine
chemical structures computationally, something like the latter is
required.  For example, graph mining algorithms can be applied to
labeled graph representations, to discover patterns in large-scale
datasets \cite{horst2009}.

@@

\subsection{Expanding Chemical Space}
\label{sec:struct-enum} 
Enumerating molecules is a combinatorial problem that has fascinated
chemists, computer scientists and mathematicians alike, for more than
a century. Indeed, many fundamental principles of graph therory and
cominatorics were developed in the context of counting isomers of
paraffin by Cayley, Polya and others. In the 1960's, Lederberg
Dejerassi and others developed algorithms to enumerate structures
based on spectral data, leading to the development of DENDRAL, widely
considered as the first expert system \cite{DENDRAL}.

From a risk reduction point of view efficient methods to enumerate
structures allows us to explore new regions of chemical space (for
example, allowing one to bypass patents), generate structures that
exhibit desired properties, identify molecules matching experimental
data and so on. It is important to realize that chemical space (the
abstract, multi-dimensional space occupied by all possible chemical
structures) is, in principle, infinite. Even considering molecules for
just 30 heavy atoms, the size of this space is on the order of
$10^{60}$ \cite{Bohacek:1996ve}. Any method that purports to enumerate
molecules will inevitably face a combinatorial explosion if
implemented na\"{i}vely.

A key application of structure enumeration is the elucidation of
structures based on spectral data \cite{Kind:2010zr}. This is
especially relevant for the identifying metabolites, small molecules
that are the byproducts of metabolic processes and which provide
insight into the biological state (diseased, fasting, etc.) of an
organism. Ideally, a chemist would gather spectral (mass, LC/MS, etc.)
and the algorithm would provide a list of structures that can give
rise to the observed spectra. Some commercial products such as MOLGEN
(\url{http://molgen.de}) are able to perform this task in a rapid
manner.

Another application of structure enumeration is in the area of Markush
structures and searches \cite{Barnard:1991vn}, where the user
specifies a pattern such as ``\emph{\ldots an aromatic ring with two
  alkyl groups attached to it}.'' Such a pattern is very general: an
alkyl group can be methyl, ethyl or any chain of $n$ carbons and there
are three possible positions for them on the ring. Even in this simple
example there are $3n^2$ possible structures. For more complex
Markushes, there can be billions of possible structures. Clearly,
explicit enumeration is not feasible. Thus one challenge is to devise
methods to search (based on structural or property similarity) through
the implicit enumerated space. A number of commerical vendors such as
Digitial Chemistry and ChemAxon provide toolkits to handle these
problems.

Structure enumeration plays a fundamental role in \emph{molecular
  design} -- the design of compounds (drugs, for instance) that
optimize some physical, chemical, or biological property or activity
\cite{Schneider:2005uq}. A key challenge in this area is to combine
enumeration algorithms with efficient property prediction and is
closely related to methods in predictive modeling of chemical
properties (Section \ref{sec:activity-mining}). This problem is also
termed \emph{inverse QSAR}, where one must devise a set of features
that describe a molecule with a specified property, and subsequently
reconstruct the molecule from those features. As a side note, the
reconstruction of molecules from feature sets is can be considered a
cryptography problem. It is relevant to the pharmaceutical industry,
as it can be necessary to share data on molecules without sharing
explicit structures. Feature sets that allow easy reconstruction of
molecular structures are thus undesirable. A number of workers have
investigated the problem of identify such one-way molecular
descriptors as well as methods to reconstruct molecules from
descriptors \cite{Masek:2008kx} and new methods that generate such
one-way molecular features but still correlate with physicochemical
properties would be invaluable.

While the complexity of molecular graph enumeration remains an open
problem, an alternative to enumerating the space is to sample it
\cite{goldberg1999}. Sampling procedures based on the Metropolis or
the Genetic Algorithms have been developed and used to elucidate
natural compounds from NMR data. It should be noted that even in the
face of the computational complexity of molecular enumeration,
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a billion chemical structures~\cite{GDB} (but only
considering molecules upto 13 heavy atoms and using only C, N, O, S
and Cl). Yet, the current enumeration software products do not
generally produce stereoisomers or tautomers; these require specific
enumeration procedures, which are still the subject of investigations
by the cheminformatics community.

Another relevant application of enumeration methods is in the
\emph{generation of chemical reaction networks}. The problem here
consists of enumerating all the possible compounds that can be
produced by applying reaction rules to a set of initial molecules. By
reversing the reaction rules one can also find sets of starting
compounds necessary for producing a given target, known as
\emph{retrosynthesis}. Designing new drugs or chemicals, understanding
the kinetics of combustion and petroleum refining, studying the
dynamics of metabolic networks, applying metabolic engineering and
synthetic biology to produce heterologous compounds in microorganisms,
all involve enumeration of reaction networks. As reviewed in Chapter
11 in \cite{faulon2010} several network enumeration techniques have
been developed. However, these techniques generally suffer from a
combinatorial explosion of product compounds. One way to limit the
number of compounds generated is to simulate the dynamics of the
network while it is being constructed and remove compounds of low
concentration. Following that idea, methods have been developed based
on the Gillespie Stochastic Simulation Algorithm (SSA) to compute
on-the-fly species concentrations. Chemical reaction network
enumeration and sampling is an active field of research, particularly
in the context of metabolism, either to study biodegradation, or to
propose metabolic engineering strategies to biosynthesize compounds of
commercial interest. The difficulty with metabolic network design is
that in addition to network generation based on reactions, one also
needs to verify that there are enzymatic events possible for enabling
reactions catalyses. That additional task requires encompassing both
chemical structure and protein sequence information and the
development of tools that are at the interface between cheminformatics
and bioinformatics.

@@

The basis of predictive modeling in cheminformatics is that the
biological activity of a molecule is a function of its chemical
structure. Together with the \emph{similar property principle}
\cite{Johnson:1990qf} (structurally similar molecules will exhibit
similar properties), the goal of any modeling approach is to capture
and characterize the correlations between structural features and the
observed biological activities. Simultaneously, such approaches must
also describe the likelihood of error when using such
models for decision making.  A variety of approaches can be employed
to assess the error in (or conversely, the reliability or confidence of) a
prediction, ranging from statistical approaches (XXX) to more
empirical approaches such as defining an \emph{applicability domain} -
the region of the input (i.e., chemical structures) that can be
reliably predicted, usually deliniated using some form of similarity
to the training set.
